---
title: "Introduction to R"
output:
  html_document: default
  pdf_document: default
  github_document: default
---


### R as a calculating environment

R can be used as a powerful calculator.
For arithmetic calculations:
```{r}
(1 + 1 / 100) ^ 100
```
```{r}
17 %% 5
```
```{r}
17 %/% 5
```
```[1]``` implies this is item 1 in a vector of output.


|symbol | meaning
|------ | --------
|+ | addition
|- | subtraction
|* | multiplication
|/ | division
|^ | exponential
|%% | modulus
|%/% | integer division


### Built-in function in R

R has a number of built in functions:  
```sin(x), cos(x), tan(x), exp(x), log(x), sqrt(x), floor(x), ceiling(x), round(x)```
```{r}
exp(1)
options(digits = 16)
exp(1)
pi
sin(pi/6)
```

```{r}
floor(2.3)
ceiling(2.3)
round(2.3)
```

### Variable

* We can assign a value to a variable and use the variable.

* For the assignment, we use command ```<-```
    * Usually, this is pronounced as "gets".

* Variable names made up of letters, numbers, . or _
    * provided it starts with a letter, or . then a letter.
    * names are case sensitive.
    * for example,
          * x, y, my_variable, a1, a2, .important_variable, x.input
    * wrong name:
          * 2016_income, .1grade, _x, y@gmail.com
    
To display the value of a variable ```x```, we type ```x``` 

* or ```print(x)``` or ```show(x)```.
```{r}
x <- 100
x
```

```{r}
print(x)
```

```{r}
show(x)
```

We can show the outcome of assignment by parentheses.
```{r}
(y <- (1 + 1 / x) ^ x)
```

When assigning, the right-hand side is evaluated first, then that value is placed in the variable on the left-hand side.
```{r}
n <- 1
n <- n + 1
n
```

R allows the use of ```=``` for variable assignment, in common with most programming languages.

The following also works.
```{r}
3 -> three
three
```
You can have multiple assignments.
```{r}
v <- w <- z <- 1
v
w
z
```


### Functions

Takes one or more argument (inputs) and produces one or more outputs (return values).

```{r}
seq(from = 1, to = 9, by =2)
```
```{r}
seq(from = 1, to = 9)
```

You can access the built-in help by ```help(function name)``` or ```?function name```:

Every function has a default order for arguments.  
If you provide arguments in this order, then they do not need to be named.
```{r}
seq(1, 9, 2)
```
```{r}
seq(to = 9, from = 1)
```
```{r}
seq(by=-2, 9, 1)
```

We will study more about functions in Chapter 4.

### Vectors

* the basic data structure in R
     * also called atomic vector
  
* indexed set of variables
* i-th element of vector ```x``` is denoted by ```x[i]```
* the types of all elements of an atomic vector should be the same
     * logical, integer, double (often called numeric), character
   
* three basic functions for constructing vectors
     * ```seq```, ```rep```, ```c```
     
* Basic three properties:
     * Type, ```typeof()```, what it is.
     * Length, ```length()```, how many elements it contains.
     * Attributes, ```attributes()```, additional arbitrary metadata.

```{r}
# short for sequence
(x <- seq(1, 20, by = 2))
```
```{r}
(x2 <- seq(1.1, 2, by = 0.1))
```

```{r}
# short for repeat
(y <- rep(3, 4))
```

```{r}
# short for combine
(z <- c(y, x))
```

* another method for sequence
```{r}
(x <- 100:110)
```
```{r}
(y <- 110:100)
```
```{r}
length(x)
```

* ```letters, LETTERS```
```{r}
# the 26 lower-case letters of the Roman alphabet
letters
```
```{r}
# the 26 upper-case letters of the Roman alphabet;
LETTERS
```


* ```seq_along()```

```seq_along(x)``` is the same as ```1:length(x)```

```{r}
x <- letters[1:5]  # a b c d e
y <- seq_along(x)
print(y)
```

* vector and index
```{r}
(x<- 100:110)
```

```{r}
# second element
x[2]
```

```{r}
# last element
x[length(x)]
```

```{r}
(x<- 100:110)

i <- c(1, 3, 2)

x[i]
```
```{r}
x[1:5]
```

minus index:
```{r}
j <- c(-1, -2, -3)
x[j]
```

but you can't mix positive and negative index, the following code is erroneous.
```{r eval=FALSE}
x[c(-1, 2)]
```

empty vetor:
```{r}
x <- c()
```

elementwise algebraic operation:
```{r}
x <- c(1, 2, 3)
y <- c(4, 5, 6)
x * y
y ^ x
```

with unequal length of vectors:
```{r}
c(1, 2, 3, 4) + c(1, 2)

2 * c(1, 2, 3)

(1:3)^2
```

This works but with warning message:
```{r eval=FALSE}
c(1, 2, 3) + c(1, 2)
```

functions taking vectors
```{r}
sqrt(1:3)

mean(1:6)

sort(c(5, 1, 3, 4, 2))
```

logical vector
```{r}
logi <- c(TRUE, FALSE, T, F)
typeof(logi)
```

character vector
```{r}
char <- c("a", "vector", "of", "characters")
typeof(char)
```

integer vector
```{r}
integ <- c(1L, 2L, 3L, 4L)
typeof(integ)
```

double vector
```{r}
doub <- c(1, 2, 3, 4)
typeof(doub)
```




#### ```typeof()```

```typeof()``` returns an (internal) object type : 

* logical - logical vector
* double - double vector
* interger - interger vector
* character - character vector
* list - see Ch.5
* builtin : R built-in function
* closure : user defined function
* and so on.

Simply speaking, the objects in R are implemented via C, 
and the internal types mean the C level data types.


#### Coercion

If we try to combine differet types of elements in a vector, they will be coerced to the most flexible type.

  * logical, integer, double, character
```{r}
c("a", 1)
```
```{r}
c(TRUE, FALSE, 0)
```


#### Example : mean and variance

compare computed mean and variance with built-in functions

```{r}
x <- c(1.2, 0.9, 0.8, 1, 1.2)

x.mean <- sum(x)/length(x)

x.mean - mean(x) 
```
```{r}
x.var <- sum((x-x.mean)^2)/(length(x)-1)
x.var - var(x)
```


#### Example : simple numerical integration

* The basic problem in numerical integration is to compute an approximate solution to a definite integral.


```{r echo=FALSE, fig.width=5, fig.height=4}
dt <- 0.005
x <- seq(0, pi/6, 0.00001)
f <- cos(x)
plot(x, f, type="l")
for(left in seq(0, pi/6, dt)){
  rect(left, 0, left+dt, cos(left))  
}
```

```{r}
dt <- 0.005
t <- seq(0, pi/6, by = dt)
ft <- cos(t)
(I <- sum(ft) * dt)
```

```t``` is a vector and ```ft``` is also a vector.

```{r}
I - sin(pi/6)
```

Note the difference between the numerical integration and theoretical value.


#### Example : exponential limit

```{r}
x <- seq(10, 200, by = 10)

y <- (1 + 1/x)^x

exp(1) - y
```
```{r fig.width=5, fig.height=4}
plot(x, y)
```

### Missing data

in R, missing data is represented by ```NA```.

```{r}
a <- NA   # assign NA to variable A
is.na(a)     # is it missing?
```
```{r}
a <- c(11, NA, 13)
is.na(a)
```
```{r}
mean(a)
```

```{r}
mean(a, na.rm = TRUE) #NAs can be removed
```

```NA``` is logical 

```{r}
typeof(NA)
```

and can be coerced to numeric, or integer, or charactor

```{r}
x <- c(NA, 10)
typeof(x[1])
```
```{r}
y <- c(NA, "abc")
typeof(y[1])
```


### Inf and -Inf

If a computation result is too big, R will return ```Inf```.

```{r}
2 ^ 1024
```
```{r}
- 2 ^ 1024
```
```{r}
1 / 0
```

### NaN

If a computation result makes little sense, then R will return ```NaN```, not a number.

```{r}
Inf - Inf
```
```{r}
0 / 0
```


### Expression and assignment

Expression is a phrase of code that can be executed.
```{r}
seq(10, 20, by=3)
```

```{r}
4
```
```{r}
mean(c(1,2,3))
```
```{r}
1 > 2
```

If the evaluation of the expression is saved using ```<-```, then it called an assignment.
```{r}
x1 <- seq(10, 20, by=3)
```
```{r}
x2 <- 1>2
```

A logical expression is formed using 

* the comparison operators 
    * ```<```, ```>```, ```<=```, ```>=```, ```==```, and ```!=``` (not equal to)

* and the logical operators 
    * ```&``` (and), ```|``` (or), and ```!``` (not).

The value of a logical expression is either ```TRUE``` or ```FALSE```.

* The integers 1 and 0 can also be used as ```TRUE``` or ```FALSE```.
```{r}
c(0, 0, 1, 1) | c(0, 1, 0, 1)
```

### x[subset]

R's subsetting operators are powerful and fast.

Subsetting is hard to learn.

We can extract a subvector using a subset as a vector of TRUE/FALSE.

```{r}
x <- 1:10
x%%4 == 0
```
```{r}
( y <- x[ x%%4==0 ] )
```

R also provide subset function, which ignore ```NA```.

* whereas ```x[subset]``` preserve ```NA```.

```{r}
x <- c(1, NA, 3, 4)
x[x > 2]
```
```{r}
subset(x, subset = x>2 )
```

For the index position of TRUE elements, use ```which(x)```
```{r}
x <- c(1, 1, 2, 3, 5, 8, 13)
which(x %% 2 == 0)
```

Explore more about subsetting:
```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```
```{r}
x[c(TRUE,FALSE,TRUE,FALSE)]
```
```{r}
x[x > 3]
```
```{r}
x[order(x)]
```

#### Example : rounding error

Many floating numbers are subject to rounding errors in digital computers.
```{r}
2 * 2 == 4
```
```{r}
sqrt(2) * sqrt(2) == 2
```
The solution is to use ```all.equal(x,y)```, which returns TRUE if the difference between x and y is smaller than some tolerance.

```{r}
all.equal(sqrt(2) * sqrt(2), 2)
```

### Sequential ```&&``` and ```||```

To evaluate ``` x && y ```, R first evaluate ```x```.
If ```x``` is ```FALSE``` then R returns ```FALSE``` without evaluating ```y```.

To evaluate ``` x || y ```, R first evaluate ```x```.
If ```x``` is ```TRUE``` then R returns ```TRUE``` without evaluating ```y```.

Sequential evalutation of ```x``` and ```y``` is useful when ```y``` is not always well defined.

```{r}
x <- 0
x == 0 || sin(1/x) == 0
```

Note that ```&&``` and ```||``` only work on scalars,
whereas ```&``` and ```|``` work on vectors elementwise.


### subsetting and assignment

Subsetting operators can be combined with assignment operator.

```{r}
x <- 1:5
x[c(1, 2)] <- 2:3
x
```
```{r}
x[-1] <- 4:1
x
```
```{r}
x[x %% 2 == 0] <- 0
x
```
```{r}
x[which(x == max(x))] <- 100
x
```



### Names

You can name a vector:
```{r}
(x <- c(a = 1, b = 2, c = 3))

```
```names``` function returns a charater vector:
```{r}
names(x)
```
```{r}
y <- 1:3
names(y) <- c("first", "second", "third")
y
```
and select elements by names:
```{r}
y["first"]
```
```{r}
y[c("first", "third")]
```




### Matrix

Matrix is created from a vector using the function matrix:

* ```matrix( data, nrow =1, ncol=1, byrow=FALSE )```
    * ```data``` : vector of length at most ```nrow*ncol```
         * if length of vector < nrow*ncol, then data is reused as many times as is needed
    * ```nrow``` : number of rows
    * ```ncol``` : number of columns
    * ```byrow = TRUE``` : fill the matrix up row-by-row
    * ```byrow = FALSE``` : fill the matrix up column-by-column, default

* ```diag(x)``` : create diagonal matrix

* ```rbind(...)``` : join matrices with rows of the same length

* ```cbind(...)``` : join matrices with columns of the same length


Example:
```{r}
(A <- matrix(1:6, nrow = 2, ncol = 3))
```
```{r}
(A <- matrix(1:6, nrow = 2, ncol = 3, byrow = TRUE))
```
```{r}
A[1, 3] <- 0
A
```

```{r}
A[, 2:3]
```

```{r}
(B <- diag(c(1, 2, 3)))
```

Matrix operation

* Usual algebraic operations, including ```*```, act elementwise. 

* To perform matrix operation, we use ```%*%```. 

* ```nrow(x)```, ```ncol(x)```    

* ```det(x)``` : determinant of ```x```    

* ```t(x)``` : transpose of ```x```    

* ```solve(A, B)``` : returns ```x``` such that ```A %*% x = B```    

* If ```A``` is invertable, the ```solve(A)``` is the inverse of ```A```.    


```{r}
A <- matrix(c(3, 5, 2, 3), nrow=2, ncol=2)
B <- matrix(c(1, 1, 0, 1), nrow=2, ncol=2)
```
```{r}
A %*% B
```
```{r}
A.inv <- solve(A)
A %*% A.inv   # we observe rounding error
```
```{r}
A^(-1)   #This is not an inverse. ^(-1) applies elementwise.
```

```cbind()``` : Combine R objects by columns

```rbind()``` : Combine R objects by rows

```{r}
(C <- cbind(A,B))
```
```{r}
(D <- rbind(A,B))
```
```{r}
colnames(C) <- c("a", "b", "c", "d")
C
```

```{r}
rownames(C) <- c("first", "second")
C
```

```{r}
# number of rows
nrow(C)
```
```{r}
# number of columns
ncol(C)
```
#### Subsetting of matrix

select columns:
```{r}
(partC1 <- C[, c(1, 2)])
```
select columns by column names:
```{r}
(partC1 <- C[, c("a", "b")])
```

```partC1``` is a matrix
```{r}
class(partC1)
```
When you select one column, it becomes a vector, not a matrix
```{r}
this_is_vector_now <- C[, "a"]
class(this_is_vector_now)
```
When you want to preserve matrix form:
```{r}
still_matrix <- C[, "a", drop = FALSE]
class(still_matrix)
```
select rows:
```{r}
(partC2 <- C[1,])
```
Or, in a combined way:
```{r}
(partC2 <- C[1, c(2, 3)])
```
```{r}
class(partC2)
```
When we use a single index, R counts the index columnwise:
```{r}
C[1]
C[2]
C[3]
C[4]
```

```{r}
is.vector(partC2)
```
```{r}
is.matrix(partC2)
```
```{r}
(Cm <- as.matrix(partC2))
```
```{r}
class(Cm)
```
```{r}
(Cv <- as.vector(Cm))
```
```{r}
class(Cv)
```


### Objects and classes

R is an object oriented language.

Every object in R is a member of a class.

You can use the ```class``` function to determine the class of object.

```{r}
# numeric vector
class(c(1, 2, 3))
```
```{r}
# character vector
class(c("c", "B", "z"))
```
```{r}
# function
class(sin)
```
```{r}
class(matrix(c(1,2)))
```


### Workspace

The objects that you create using R remain in existence until you explicitly delete them.

* ```rm(x)``` : remove object ```x```

* ```rm(list=ls())``` : remove all objects


### Working directory

When you run R, it uses one of the directories on your hard drive as a working directory,

* where it looks for user-written programs and data files.

Check the working directory.
```{r eval=FALSE}
getwd()
```

Change the working directory to "dir"

```{r eval=FALSE}
setwd("dir")
```

```/``` is for directory and file address, ```.``` refers current directory, ```..``` refers parent directory

### Writing script

We can type and evaluate all possible R expression at the prompt, it is much more convenient to write scripts,

* which simply comprise collections of R expression.

* We use the terms program and code synonymously with script.

You can use built-in editor in Rgui or Rstudio.    

### Help

To find out more about an R command or function x, you can type ```help(x)``` or just ```?x```.

If you cannot remember the exact name, then ```help.search("x")```.

HTML help command : ```help.start()```

### package

R provides various useful packages to help you.

https://cran.r-project.org/web/packages/

To install a package:  

```install.packages("packagename")```  

To access the package:  

```library("packagename")```

Or use package menu.

R documentation : https://www.rdocumentation.org/
