---
title: "R Basic Programming"
output:
  html_document: default
  pdf_document: default
  github_document: default
---




### Branching with if

```{r eval=FALSE}
if ( logical_expression ) {  
  expression_1
}  
```

```{r eval=FALSE}
if (logical_expression) {   
  expression_1
} else {
  expression_2
}  
```

If ```logical_expression``` is ```TRUE``` then the first group of expression is executed and the second group of expression is not executed.

Conversely if ```logical_expression``` is ```FALSE``` then only the second group expression is executed.


If you type:   

```{r eval=FALSE}
if (logical_expression) {
  expression_1
}
else {  
  expression_2  
}  
```

then you get an error.

This is because R is believes the if statement is finished before it sees the else part.  
That is, R treats the else as the start of a new command, but there is no command that starts with an else.

The followings are equivalent.


```{r eval=FALSE}
if (logical_expression_1){    
  expression_1    
} else {   
    if (logical_expression_2){   
      expression_2  
    } else {   
      expression_3  
    }      
}
```
```{r eval=FALSE}
if (logical_expression_1){   
  expression_1   
} else if (logical_expression_2){   
  expression_2   
} else {   
  expression_3   
}
```

#### Example : root of quadratic

```{r}
rm(list=ls())

a0 <- 1; a1 <- 6; a2 <- 7
discrim <- a1^2 - 4*a2*a0

if (discrim > 0) {
  roots <- c( (-a1 + sqrt(a1^2 - 4*a2*a0))/(2*a2), (-a1 - sqrt(a1^2 - 4*a2*a0))/(2*a2) )
} else if (discrim == 0) {
  roots <- -a1/(2*a2)
} else {
  roots <- c() 
}

show(roots)
```

### Looping with for

```{r eval=FALSE}
for (x in vector) {  
  expression_1   
  ...   
}
```

where ```x``` is a simple variable.

#### Example : summing a vector

We have a built-in function sum() but for an illustrative purpose:

```{r}
(x_list <- seq(1, 9, by = 2))
sum_x <- 0
for (x in x_list) {
  sum_x <- sum_x + x
  cat("The current loop element is", x, "\n")
  cat("The cumulative total is", sum_x, "\n")
}
```
* ```cat()``` : for concatenate
* ```\n``` : new line

#### Example : n factorial

```{r}
n <- 6
n_factorial <- 1
for (i in 1:n) {
  n_factorial <- n_factorial * i
}
show(n_factorial)
```



#### Example : pension	

```{r fig.width=5, fig.height=4}
r <- 0.11             # Annual interest rate
term <- 10            # Forecast duration (in years)
period <- 1/12        # Time between payments (in years)
payments <- 100       # Amount deposited each period
# Calculations

n <- floor(term/period)  # Number of payments
pension <- 0
for (i in 1:n) {
  pension[i+1] <- pension[i]*(1 + r*period) + payments
}
time <- (0:n)*period

# Output
plot(time, pension)

```

#### Example : redimensioning


Program 1 and 2 produce the same result but 1 is faster.
```{r eval=FALSE}
#Program1
n <- 100000
x <- rep(0, n)
for (i in 1:n) x[i] <- i
```
```{r eval=FALSE}
#Program2
n <- 100000
x <- 1
for (i in 2:n) x[i] <- i
```
Program1 of preallocation is faster than Program2 of redimensioning.

### Looping with while

```{r eval=FALSE}
while (logical_expression) {   
  expression_1   
  ...   
}
```
* ```logical_expression``` is evaluated first.
* If it is ```TRUE```, then the expression in ```{ }``` is executed.
  * back to the start of the command.
* If ```logical_expression``` is ```FALSE```, the loop stop.

* We can always rewrite a for loop as a while loop.


#### Example : Fibonacci numbers

```{r}
rm(list=ls())
F <- c(1, 1)
n <- 2
while (F[n] <= 100) {
  n <- n +1
  F[n] <- F[n-1] + F[n-2]
}
cat("The first Fibonacci number > 100 is F(", n, ") =", F[n], "\n")
```

### Example : compound interest rate

```{r}
rm(list=ls())
r <- 0.11                # Annual rate
period <- 1/12           # Time between repayments (in years)
debt_initial <- 1000     # Amount borrowed
repayments <- 12         # Amount repaid each period

# Calculations
time <- 0
debt <- debt_initial
while (debt > 0) {
  time <- time + period
  debt <- debt*(1 + r*period) - repayments
}

# Output
cat('Loan will be repaid in', time, 'years\n')
```

### Vector-based programming

Using vector operations is more efficient computationally.  
Sum of the first n squares using loop:
```{r}
n <- 100; S <- 0
for (i in 1:n)  S <- S + i^2
print(S)
```

Alternatively, using vector operations:
```{r}
sum((1:n)^2)
```

#### ```ifelse```

The ```ifelse``` function performs elementwise conditional evaluation upon a vector.  
```ifelse``` returns a value with the same shape as test which is filled with elements selected from either yes or no depending on whether the element of test is ```TRUE``` or ```FALSE```.

* ```ifelse(test, A, B)```
  * ```test``` : logical expression
  * returns a vector consist of
    * ```A``` : when element of test are ```TRUE```
    * ```B``` : when element of test are ```FALSE```

```{r}
x <- c(-2, -1, 1, 2)
ifelse( x>0, "Positive", "Negative")
```

#### ```pmin``` and ```pmax```
* vectorised versions of the minimum and maximum  
* ```pmax``` and ```pmin()``` take one or more vectors as arguments, recycle them to common length and return a single vector giving the parallel maxima (or minima) of the argument vectors.

```{r}
pmin( c(1, 2, 3), c(3, 2, 1), c(2, 2, 2))
```


### Program flow

```{r}
x <- 3
for (i in 1:3) {
  show(x)
  if (x %% 2 == 0) {
    x <- x/2
  } else {
    x <- 3*x + 1
  }
}
show(x)
```


### Basic debugging : correcting errors

* To find an error or bug, you need to see how your variables change.
  * include statements like ```cat("var=", var, "\n")```

* dry run : using simple starting conditions for which you know what the answer should be.
  * use short and simple versions of the final program

* Use graph and summary statistics.

* Careful use of indentation.

#### Example

```{r}
x <- 3
for (i in 1:3) {
  show(x)
  cat("i = ", i, "\n")
  if (x %% 2 == 0) {
    x <- x/2
  } else {
    x <- 3*x + 1
  }
}
show(x)

```


### Good programming habits

* Good programming is clear rather than clever.
  * in practice, much more time is spent correcting and modifying codes than is ever spent writing them.
* Write comments.
* Variable name should be descriptive.
  * Avoid using reserved names of functions
    * for example, ```t```, ```c``` and ```q``` are all function names in R
    * check with ```exists()``` function
* Use black line to separate sections.
