---
title: "Programming with functions"
output:
  html_document: default
  github_document: default
  pdf_document: default
---

### Functions

Functions are a fundamental building block of R.

In R, the operations that do all the work are functions.

A function is an object in R that takes some input objects (called arguments) and returns an ouput object.

A function has a form:
```{r eval=FALSE}
function_name <- function(argument_1, argument_2, ...) {
  expression_1
  expression_2
  ...
  return(output)
}
```
* ```argument_1, arguments_2``` are the names of variables as inputs for the function.
* ```function_name``` is the name of function. You change ```function_name``` as you want.

To call or run the function we type:
```{r eval=FALSE}
function_name(x1, x2, ...)
```
the value of this expression is the value of output.

* A function may have more than one ```return``` statement, in which case it stops after executing the first one it reaches.
* If there is no statement ```return(output)``` then the value returned by the function is the value of the last expression in the braces.

* A function always returns a value.
  * For some functions the value returned is unimportant.
  * In such cases one usually omits the return statement, or returns NULL.
* If the value returned by a function is not assigned to a variable, then it is printed.

* The most important advantage of using a function is that once it is loaded, it can be used again and again without having to reload it.

* The second most important use of functions is to break down a programming task into smaller logical units.


#### Example : Find zeros of ```a2*x^2+a1*x+a0=0```

```{r}
quad3 <- function(a0, a1, a2) {
  if(a2 == 0 && a1 == 0 && a0 == 0) {
    roots <- NA
  } else if (a2 == 0 && a1 == 0 ) {
    roots <- NULL
  } else if ( a2 == 0 ) {
    roots <- -a0/a1
  } else {
    discrim <- a1^2 - 4*a2*a0
    if (discrim > 0) {
      roots <- (-a1 +c(1,-1) * sqrt(a1^2-4*a2*a0))/(2*a2)
    } else if (discrim == 0) {
      roots <- -a1/(2*a2)
    } else {
      roots <- NULL
    }
  }
  roots
}
```

```{r}
quad3(6, -5, 1)
```


#### Example : n choose r

The number of ways that you can choose r things from a set of n, ignoring the order, is  

$$ \frac{n!}{r!(n-r)!} $$
```{r}
n_factorial <-function(n) {
  n_fact <- prod(1:n)
  return(n_fact)
}
n_choose_r <- function(n, r) {
  n_ch_r <- n_factorial(n) / n_factorial(r) / n_factorial(n-r)
  return(n_ch_r)
}
```
```{r}
n_choose_r(10,6)
```
```{r}
n_choose_r(10,4)
```

#### Example : Winsorised mean
k-th Winsorised mean of ordered data $x=\{ x_1, ..., x_n \}$ is defined as  

$$ w_k = \frac{(k+1)x_{k+1} + x_{k+2} + \cdots + x_{n-k+1} + (k+1) x_{n-k} }{n} $$
```{r}
wmean <- function(x, k) {
  x <- sort(x)
  n <- length(x)
  x[1:k] <- x[k + 1]
  x[(n - k + 1):n] <- x[n - k]
  mean(x)
}
```


#### Exmple : Swap

swap values of ```x[1]``` and ```x[2]```
```{r eval=FALSE}
swap <- function(x){
  y <- x[2]
  x[2] <- x[1]
  x[1] <- y
  return(x)
}
x <- c(7, 8, 9)
x[1:2] <- swap(x[1:2])
x[2:3] <- swap(x[2:3])
```


### Scope and its consequences

* Argument and variables defined within a function exist only within that function.
   * If you define and use a variable x inside a function, it does not exist outside the function.
   * If variables with the same name exist inside and outside a function, then they are separate and do not interact at all.
* The variable defined outside the function can be seen inside the function (provided there is not a variable with the same name defined inside).

```{r}
test <- function(x) {
  y <- x+1
  return(y)
}
test(1)
```
```{r eval=FALSE}
y
### Error : Object "y" not found
```
```{r}
test2 <- function(x) {
  y <- x + z
  return(y)
}
z <- 1
test2(1)
z <- 2
test2(1)
```

### Optional argument and default value

* To give ```argument_1``` the default value ```x1``` we use ```argument_1 = x1``` within the function definition.
* If an argument has a default, then it may be omitted when calling the function, in which case the default is used.


```{r}
test3 <- function(x = 1, y = 1, z = 1) {
  return(x * 100 + y * 10 + z)
}
test3(2, 2)
test3(y = 2, z = 2)
```


### Every operation in R is function call

All work in R is done by functions.
For example,
```{r}
k <- 3
k
```
This function does similar operation with above.
```{r}
# `<-` is a function name
`<-`(k, 4)    # k <- 4
k
```
In addition, if we have
```{r}
animals <- c("cow", "dog", "cat", "pig")
```
and you want 
```{r}
animals[3]
```
This is equivalent to
```{r}
`[`(animals, 3)
```
In the above ``` `[` ``` is a function name.

Similarly,
```{r eval=FALSE}
animals[4] <- "duck"
```
can be achieved by
```{r}
(animals <- `[<-`(animals, 4, "duck"))
```
In practice, we do not use the above expression, but it is useful to know that every operation in R is a function.

Other example:
```{r}
x <- 3
y <- 4
x + y
```
```{r}
# functional form of arithmetic expression, which is equal to x + y
`+`(x, y)
```

```{r}
if (x > y) print("x is larger than y") else print("y is larger than x")
```

```{r}
# functional form of if-then expression
`if`(x > y, print("x is larger than y"), print("y is larger than x"))
```

### Function as an object

A function in R is just another object that is assigned to a symbol.

```{r}
f <- function(x, y) c(x + 1, y + 2)
f(1,2)
```

You can type the name of function to see the code for it.

```{r}
f
```
### Vector-based programming

* Many R functions are vectorised.
* To further facilitate vector-based programming, R provides functions that enable the vectorisation of user-defined functions.
* ```apply```, ```sapply```, ```lapply```, ```tapply```, and ```mapply```.
* ```sapply(X, FUN)```
   * apply function ```FUN``` to every element of vector ```X```.




#### Exmaple : Simple sapply

```{r}
sapply(1:5, sqrt)
```
which is equivalent to
```{r}
sqrt(1:5)
```


#### Example : Simple sapply with anonymous function

Anonymous function is a function without a name.
You use an anonymous function when it is not worth the effort to give it a name.

```{r}
(function(x) x^2 + 1)(10)
```

Anonymous function can be applied to ```sapply```:
```{r}
sapply(1:10, function(x) x^2 + 1)
```

```{r}
sapply(1:4, function(x) paste0("a", x))
```

```{r}
m <- matrix(1:9, nrow = 3, byrow = T)
sapply(1:3, function(x) sum(m[x, ]))   # equivalent to rowSums(m)
```

<!-- #region -->
#### Example : Density of primes with sapply
 
* Let $\rho(n)$ be the number of primes less than or equal to n
* Then

$$ \lim_{n \rightarrow \infty} \frac{\rho(n)\log(n)}{n} = 1$$


* To check this, first we define a function ```prime``` that tests if a given integer is prime.

* We then use ```sapply``` to apply ```prime``` to the vector ```2:n```.
<!-- #endregion -->

```{r}
prime <- function(n) {
  if (n == 1) {
    is.prime <- FALSE
  } else if (n == 2) {
    is.prime <- TRUE
  } else {
    is.prime <- TRUE
    for (m in 2:(n/2)) {
      if ( n%%m == 0 ) is.prime <- FALSE
    }
  }
  return(is.prime)
}
```
```{r}
x <- 1:22
y <- sapply(x, prime)
tapply(x, y, max)
```

```{r}
n <- 1000
m.vec <- 2:n
primes <- sapply(m.vec, prime)
num.primes <- cumsum(primes)
plot(m.vec, num.primes/m.vec, type = "l", main = "prime density", xlab = "n", ylab = "")
lines(m.vec, 1/log(m.vec), col= "red")
```

### Recursive programing

* A recursive program is one that calls itself.
* This is useful because many algorithms are recursive in nature.

#### n factorial

```{r}
nfact2 <- function(n) {
  if (n==1) {
    cat("called nfact2(1)\n")
    return(1)
  } else {
    cat("called nfact2(", n, ")\n", sep="")
    return(n*nfact2(n-1))
  }
}
```


#### loop to recursive function

Any loop can be converted to a recursive function.

```{r}
total <- 0
n <- 10
for (i in 1:n) total <- total + i
total
```

```{r}
sum_n <- function(n){
  if (n == 1) 1
  else n + sum_n(n - 1)
}
sum_n(10)
```


```<<-``` can change the values in the parent environment.
Using ```<<-```, another version of recursive function can be constructed.

```{r}
total <- 0
sum_n <- function(n){
  total <<- total + n
  if(n > 1) sum_n(n - 1)
}
sum_n(10)
total
```


#### Sieve of Eratosthenes
The Sieve of Eratosthenes is an algorithm for finding all of the primes less than or equal to a given number $n$.

1. Start with the list $2,3,\cdots,n$ and $p=2$.
2. Remove from the list all elements that are multiples of $p$ (but keep $p$ itself).
3. Increase $p$ to the smallest element of the remaining list that is larger than the current $p$.
4. If $p$ is larger than sqrt of $n$ then stop, otherwise go back to step 2.

```{r}
primesieve <- function(sieved, unsieved) {
  p <- unsieved[1]
  n <- unsieved[length(unsieved)]
  if ( p^2 > n ) {
    return(c(sieved, unsieved))
  } else {
    unsieved <- unsieved[unsieved %% p != 0]
    sieved <- c(sieved, p)
    return(primesieve(sieved, unsieved))
  }
}

```





